---
output: github_document
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  fig.retina = 2,
  fig.align = 'center'
)
```

# kalaiwash

<!-- badges: start -->

[![License: CC BY
4.0](https://img.shields.io/badge/License-CC_BY_4.0-lightgrey.svg)](https://creativecommons.org/licenses/by/4.0/)

<!-- badges: end -->

The goal of kalaiwash is to provide the baseline and endline household
survey data of the KALAI water, sanitation and hygiene project of
HELVETAS in Nampula Province, Mozambique. The project was funded by
charity: water. Baseline data were collected in November 2024 and
endline data in June 2025 in the districts of Larde, Memba, Moma and
Mecuburi. The data describe drinking water sources classified with the
WHO/UNICEF Joint Monitoring Programme (JMP) service ladder, water
collection times and volumes, sanitation and handwashing practices, and
household water insecurity measured with the 12-item Household Water
Insecurity Experiences (HWISE) Scale.

## Installation

You can install the development version of kalaiwash from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("openwashdata/kalaiwash")
```

```{r}
## Run the following code in console if you don't have the packages
## install.packages(c("dplyr", "knitr", "readr", "stringr", "gt", "kableExtra"))
library(dplyr)
library(knitr)
library(readr)
library(stringr)
library(gt)
library(kableExtra)
```

Alternatively, you can download the individual datasets as a CSV or XLSX
file from the table below.

1.  Click Download CSV. A window opens that displays the CSV in
    your browser.
2.  Right-click anywhere inside the window and select "Save Page As...".
3.  Save the file in a folder of your choice.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

extdata_path <- "https://github.com/openwashdata/kalaiwash/raw/main/inst/extdata/"

read_csv("data-raw/dictionary.csv") |> 
  distinct(file_name) |> 
  dplyr::mutate(file_name = str_remove(file_name, ".rda")) |> 
  dplyr::rename(dataset = file_name) |> 
  mutate(
    CSV = paste0("[Download CSV](", extdata_path, dataset, ".csv)"),
    XLSX = paste0("[Download XLSX](", extdata_path, dataset, ".xlsx)")
  ) |> 
  knitr::kable()

```

## Data

The package provides access to one dataset.

```{r}
library(kalaiwash)
```

### kalaiwash

The dataset `kalaiwash` contains one row per household interview. It
has `r nrow(kalaiwash)` observations and `r ncol(kalaiwash)` variables.

```{r}
kalaiwash |> 
  head(3) |> 
  gt::gt() |>
  gt::as_raw_html()
```

For an overview of the variable names, see the following table. The
`options` column lists the levels of each categorical variable.

```{r echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv("data-raw/dictionary.csv") |>
  dplyr::filter(file_name == "kalaiwash.rda") |>
  dplyr::select(variable_name, variable_type, description, options) |> 
  knitr::kable() |> 
  kableExtra::kable_styling("striped") |> 
  kableExtra::scroll_box(height = "200px")
```

Variables that were calculated in the original Excel workbook
(`jmp_improved`, `jmp_water_service`, `total_collect_time`,
`total_liters`, `liters_person`, `hwise_score` and
`hwise_insecurity_level`) were recalculated in R from the raw survey
responses and verified against the Excel values. The 12 HWISE items are ordered
factors whose levels refer to the number of times in the last four
weeks: Never (0), Rarely (1–2), Sometimes (3–10), Often (11–20) and
Always (more than 20). The HWISE score sums
the 12 items scored Never 0, Rarely 1, Sometimes 2, Often or Always 3.
The insecurity levels follow Frongillo et al. (2024): No-to-marginal
(0–2), Low (3–11), Moderate (12–23) and High (24–36).

## Example

The example compares the JMP drinking water service level of households
between the baseline and the endline survey.

```{r}
library(kalaiwash)
library(dplyr)

kalaiwash |> 
  count(survey_type, jmp_water_service) |> 
  group_by(survey_type) |> 
  mutate(share = round(100 * n / sum(n))) |> 
  ungroup() |> 
  knitr::kable(col.names = c("Survey", "JMP service level", "Households", "Share (%)"))
```

The share of households with at least basic drinking water service and
the distribution of HWISE insecurity levels can be compared in the same
way.

```{r}
kalaiwash |> 
  count(survey_type, hwise_insecurity_level) |> 
  group_by(survey_type) |> 
  mutate(share = round(100 * n / sum(n))) |> 
  ungroup() |> 
  knitr::kable(col.names = c("Survey", "HWISE insecurity level", "Households", "Share (%)"))
```

## License

Data are available as
[CC-BY](https://github.com/openwashdata/kalaiwash/blob/main/LICENSE.md).

## Citation

Please cite this package using:

```{r}
citation("kalaiwash")
```

## References

Young, S. L., Boateng, G. O., Jamaluddine, Z., et al. (2019). The
Household Water InSecurity Experiences (HWISE) Scale: development and
validation of a household water insecurity measure for low-income and
middle-income countries. *BMJ Global Health*, 4(5), e001750.
<https://doi.org/10.1136/bmjgh-2019-001750>

Frongillo, E. A., Bethancourt, H. J., Miller, J. D., Young, S. L., &
the HWISE Research Coordination Network (2024). Identifying ordinal
categories for the Water Insecurity Experiences Scales. *Journal of
Water, Sanitation and Hygiene for Development*, 14(11), 1066–1078.
<https://doi.org/10.2166/washdev.2024.042>
