---
output: github_document
always_allow_html: true
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  fig.retina = 2,
  fig.align = 'center'
)
```

# kalaiwash

<!-- badges: start -->

[![License: CC BY
4.0](https://img.shields.io/badge/License-CC_BY_4.0-lightgrey.svg)](https://creativecommons.org/licenses/by/4.0/)
[![R-CMD-check](https://github.com/openwashdata/kalaiwash/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/openwashdata/kalaiwash/actions/workflows/R-CMD-check.yaml)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.23164217.svg)](https://zenodo.org/doi/10.5281/zenodo.23164217)
<!-- badges: end -->

The goal of kalaiwash is to provide the baseline and endline household
survey data of the KALAI water, sanitation and hygiene project of
HELVETAS in Nampula Province, Mozambique. The project was funded by
charity. Baseline data were collected in November 2024 and
endline data in June 2025 in the districts of Larde, Memba, Moma and
Mecuburi. The data describe drinking water sources classified with the
WHO/UNICEF Joint Monitoring Programme (JMP) service ladder, and
household water insecurity measured with the 12-item Household Water
Insecurity Experiences (HWISE) Scale.

## Installation

You can install the development version of kalaiwash from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("openwashdata/kalaiwash")
```

```{r packages, include = FALSE}
## Run the following code in console if you don't have the packages
## needed to compile the README
## install.packages(c("dplyr", "tidyr", "ggplot2", "patchwork", "knitr", "readr", "stringr", "kableExtra"))
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(knitr)
library(readr)
library(stringr)
library(kableExtra)
```

Alternatively, you can download the individual datasets as a CSV or XLSX
file from the table below.

1.  Click Download CSV. A window opens that displays the CSV in
    your browser.
2.  Right-click anywhere inside the window and select "Save Page As...".
3.  Save the file in a folder of your choice.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

extdata_path <- "https://github.com/openwashdata/kalaiwash/raw/main/inst/extdata/"

read_csv("data-raw/dictionary.csv") |> 
  distinct(file_name) |> 
  dplyr::mutate(file_name = str_remove(file_name, ".rda")) |> 
  dplyr::rename(dataset = file_name) |> 
  mutate(
    CSV = paste0("[Download CSV](", extdata_path, dataset, ".csv)"),
    XLSX = paste0("[Download XLSX](", extdata_path, dataset, ".xlsx)")
  ) |> 
  knitr::kable()

```

## Data

The package provides access to one dataset.

```{r}
library(kalaiwash)
```

### kalaiwash

For an overview of the variable names, see the following table. The
`options` column lists the levels of each categorical variable.

```{r echo=FALSE, message=FALSE, warning=FALSE}
readr::read_csv("data-raw/dictionary.csv") |>
  dplyr::filter(file_name == "kalaiwash.rda") |>
  dplyr::select(variable_name, variable_type, description, options) |> 
  knitr::kable() |> 
  kableExtra::kable_styling("striped") |> 
  kableExtra::scroll_box(height = "200px")
```


## Example

The example compares the JMP drinking water service level of households
between the baseline and the endline survey.

```{r example-jmp, echo = FALSE}
## Run the following code in console if you don't have the packages
## needed to run the example
## install.packages(c("dplyr", "tidyr", "knitr", "ggplot2", "patchwork"))
library(kalaiwash)
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)
library(patchwork)

kalaiwash |> 
  count(survey_type, jmp_water_service, .drop = FALSE) |> 
  group_by(survey_type) |> 
  mutate(share = round(100 * n / sum(n))) |> 
  ungroup() |> 
  select(-n) |> 
  pivot_wider(names_from = survey_type, values_from = share) |> 
  arrange(jmp_water_service) |> 
  kable(col.names = c("JMP service level", "Share baseline (%)", "Share endline (%)"))
```

<details>
<summary>Show the code</summary>

```{r example-jmp-code, ref.label = "example-jmp", eval = FALSE}
```

</details>

The share of households with at least basic drinking water service and
the distribution of HWISE insecurity levels can be compared in the same
way.

```{r example-hwise, echo = FALSE}
kalaiwash |> 
  count(survey_type, hwise_insecurity_level, .drop = FALSE) |> 
  group_by(survey_type) |> 
  mutate(share = round(100 * n / sum(n))) |> 
  ungroup() |> 
  select(-n) |> 
  pivot_wider(names_from = survey_type, values_from = share) |> 
  arrange(hwise_insecurity_level) |> 
  kable(col.names = c("HWISE insecurity level", "Share baseline (%)", "Share endline (%)"))
```

<details>
<summary>Show the code</summary>

```{r example-hwise-code, ref.label = "example-hwise", eval = FALSE}
```

</details>

Figure 1 was taken from [Advancing standard WASH metrics with experiential 
indicators in Mozambique](https://github.com/ds4owd-002/project-johnbrogan-alt). 
The left panel shows the change in the JMP drinking water service level and the 
right panel the change in the HWISE water insecurity level between the baseline 
and the endline survey. Each band connects the share of households in a level at
baseline with the share at endline.

```{r figure-1, echo = FALSE, fig.width = 9, fig.height = 5, fig.cap = "Figure 1: Changes in drinking water service and water insecurity experience between baseline (November 2024) and endline (June 2025)."}
# share of households per level and survey round, with the stacking
# positions of each band (best level at the bottom)
level_shares <- function(data, level) {
  data |>
    count(survey_type, level = {{ level }}) |>
    group_by(survey_type) |>
    mutate(percent = 100 * n / sum(n)) |>
    ungroup() |>
    complete(survey_type, level, fill = list(n = 0, percent = 0)) |>
    mutate(level = factor(level, levels = rev(levels(level)))) |>
    arrange(survey_type, level) |>
    group_by(survey_type) |>
    mutate(ymax = cumsum(percent), ymin = ymax - percent,
           ymid = (ymin + ymax) / 2, x = as.integer(survey_type)) |>
    ungroup()
}

# four corners of each band: baseline bottom, baseline top, endline top,
# endline bottom
level_bands <- function(shares) {
  shares |>
    select(level, x, ymin, ymax) |>
    pivot_longer(c(ymin, ymax), names_to = "edge", values_to = "y") |>
    arrange(level, x, if_else(x == 1, y, -y))
}

level_plot <- function(shares, colours, title, legend_title) {
  n_surveys <- shares |>
    group_by(survey_type) |>
    summarise(n = sum(n), .groups = "drop")
  labels <- shares |>
    filter(percent > 0) |>
    mutate(label = paste0(round(percent), "%"),
           x = if_else(x == 1, x - 0.04, x + 0.04),
           hjust = if_else(x < 1, 1, 0))

  ggplot() +
    geom_polygon(data = level_bands(shares),
                 aes(x, y, group = level, fill = level),
                 colour = "grey20", linewidth = 0.3) +
    geom_text(data = labels, aes(x, ymid, label = label, hjust = hjust),
              size = 3.2, colour = "grey20") +
    scale_fill_manual(values = colours, name = legend_title) +
    scale_x_continuous(
      breaks = 1:2, limits = c(0.65, 2.35),
      labels = paste0(n_surveys$survey_type, "\nn = ", n_surveys$n)
    ) +
    scale_y_continuous(breaks = seq(0, 100, 25), expand = expansion(c(0, 0.02))) +
    labs(x = NULL, y = "Households (%)", title = title) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.major.x = element_blank(),
          panel.grid.minor = element_blank(),
          plot.title = element_text(size = 12, hjust = 0.5))
}

p_jmp <- kalaiwash |>
  level_shares(jmp_water_service) |>
  level_plot(
    colours = c("Basic" = "#00B8EC", "Limited" = "#FFF59D",
                "Unimproved" = "#FFB300", "Surface water" = "#E65100"),
    title = "Drinking water", legend_title = "JMP service level"
  ) +
  theme(legend.position = "left")

p_hwise <- kalaiwash |>
  level_shares(hwise_insecurity_level) |>
  level_plot(
    colours = c("No-to-marginal" = "#0066A6", "Low" = "#00B8EC",
                "Moderate" = "#FFB300", "High" = "#EF414A"),
    title = "Water insecurity", legend_title = "HWISE level"
  ) +
  labs(y = NULL)

p_jmp + p_hwise
```

<details>
<summary>Show the code</summary>

```{r figure-1-code, ref.label = "figure-1", eval = FALSE}
```

</details>

## License

Data are available as
[CC-BY](https://github.com/openwashdata/kalaiwash/blob/main/LICENSE.md).

## Citation

Please cite this package using:

```{r}
citation("kalaiwash")
```

## References

Young, S. L., Boateng, G. O., Jamaluddine, Z., et al. (2019). The
Household Water InSecurity Experiences (HWISE) Scale: development and
validation of a household water insecurity measure for low-income and
middle-income countries. *BMJ Global Health*, 4(5), e001750.
<https://doi.org/10.1136/bmjgh-2019-001750>

Frongillo, E. A., Bethancourt, H. J., Miller, J. D., Young, S. L., &
the HWISE Research Coordination Network (2024). Identifying ordinal
categories for the Water Insecurity Experiences Scales. *Journal of
Water, Sanitation and Hygiene for Development*, 14(11), 1066–1078.
<https://doi.org/10.2166/washdev.2024.042>
